Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-30 optim.qmd
lots of small typos fixed and to the best of our knowledge statement removed

The primary goal of the study reported here is to investigate the performance of JSO in PP optimisation for the guided tour. It is of interest to assess how quickly and closely the optimiser reaches a gobal optima, for various PP indexes that may have differing complexities. To observe the performance of JSO with different types of PP indexes, metrics are introduced to capture specific properties of the index including squintability and smoothness as introduced in @laa_using_2020. Metrics for squintability and smoothness are precisely defined here, something that is new for the field.  A series of simulation experiments using specially structured datasets and various PP indexes are conducted to assess the behaviour of JSO under different settings and its sensitivity to hyper-parameter choices.  The relationship between the JSO performance, hyper-parameter tuning and properties of PP
indexes is analysed to provide guidance for practitioners using projection pursuit, and developers of new indexes.

@@

where $\nu>0$ is the smoothness parameter and where $\mathcal{K}_\nu$ is
the modified Bessel function. The Matérn covariance function can be
expressed analytically when $\nu$ is a half-integer, the most popular
values in the literature being $1/2$, $3/2$ and $5/2$ . The parameter
$\nu$, called *smoothness parameter*, controls the decay of the covariance
function. As such, it is an appropriate measure of smoothness of a
random field.

@@

In this subsection, a metric to capture the notion of squintability is proposed, and two approaches to compute this metric numerically are detailed.

@@

which is a decreasing logistic function depending on two parameters
$\theta_2$ and $\theta_3$ , such that $\ell(\theta_{2})=\frac{1}{2}$.
Then, define

@@

which depends on three additional parameters, $\theta_1$, $\theta_2$ ,
and $x_{\max}$, such that $f(0)=\theta_1$ and $f(x_{\max})=\theta_4$.
Under the parametric model (@eq-parametric), the squintability metric
(@eq-squintability) can be shown to be equal to

@@

To the best of our knowledge, this is the first attempt to measure mathematically the
notion of squintability.

@@

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: Parameters estimated from the Gaussian process (including variance, range, smoothness, and nugget) and scaled logistic function ($\theta_1$ to $\theta_4$) for the pipe-finding and sine-wave finding problems.  The column "smooth" and "squint" represent the smoothness and squintability measures. 
dt <- tibble(shape = c(rep("pipe", 4), rep("sine", 8)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  mutate(index = ifelse(index == "dcor2d_2", "dcor2d", index)) 
dt |> 
  knitr::kable(
    digits = 3, format = "latex", 
    col.names = c(
      "shape", "index", "d",
      "variance", "range", "smooth", "nugget",
      "$\\theta_1$", "$\\theta_2$", "$\\theta_3$", "$\\theta_4$", "squint"),
    booktabs = T, escape = FALSE, row.names = TRUE) |> 
  column_spec(1, border_left = TRUE) |> 
  column_spec(c(4, 8, 13), border_right = TRUE) |> 
  kable_styling(font_size = 7) 
```

```{r}
#| label: tbl-mod-data
#| tbl-cap: The first 7 rows of the datasets processed for modelling. 
sim_df |> 
  dplyr::select(index, d, smoothness, squintability, n_jellies, max_tries, 
                -I_max, P_J, time) |> 
  arrange(index, d, n_jellies, max_tries) |> 
  head(7) |> 
  kable(digits = 3,
        col.names = c("Index", "D", "Smoothness", "Squintability", 
                      "Num. of jellyfish", "Max. Num. of tries", 
                      "Success rate", "Time"), 
        booktabs = T) |> 
  kable_styling(latex_options = "hold_position", font_size = 7)
```

```{r}
#| label: tbl-mod-output
#| tbl-cap: Model estimates of proportion of jellyfish success on optimisation properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function. The variable smoothness, squintability, number of jellyfish and maximum number of tries are positively associated with JSO success rate while data dimension and being flagged as long runtime are negatively associated with the success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.
broom::tidy(mod1) |> 
  mutate(term = c("Intercept", "Smoothness", "Squintability", "dimension (d)", 
                  "long time", "number of jellyfish", "maximum number of tries")) |>
  kable(digits = 3) |> 
  column_spec(c(1,5), border_right = TRUE) |> 
  column_spec(1, border_left = TRUE) |> 
  kable_styling(font_size = 7) 
```

>>>>

The primary goal of the study reported here is to investigate the performance of JSO in PP optimisation for the guided tour. It is of interest to assess how quickly and closely the optimiser reaches a gobal optima, for various PP indexes that may have differing complexities. To observe the performance of JSO with different types of PP indexes, metrics are introduced to capture specific properties of the index including squintability (based on @barnett1981interpreting's squint angle) and smoothness as introduced in @laa_using_2020. Here we mathematically define metrics for squintability and smoothness, which is new for the field.  A series of simulation experiments using specially structured datasets and various PP indexes are conducted to assess the behaviour of JSO under different settings and its sensitivity to hyper-parameter choices.  The relationship between the JSO performance, hyper-parameter tuning and properties of PP
indexes is analysed to provide guidance for practitioners using projection pursuit, and developers of new indexes.

@@

where $\nu>0$ is the smoothness parameter and where $\mathcal{K}_\nu$ is
the modified Bessel function. The Matérn covariance function can be
expressed analytically when $\nu$ is a half-integer, the most popular
values in the literature being $1/2$, $3/2$ and $5/2$. The parameter
$\nu$, called *smoothness parameter*, controls the decay of the covariance
function. As such, it is an appropriate measure of smoothness of a
random field.

@@

Squintability has been a useful concept. Here, it is defined mathematically and two approaches to compute this metric numerically, are described.

@@

which is a decreasing logistic function depending on two parameters $\theta_2$ and $\theta_3$, such that $\ell(\theta_{2})=\frac{1}{2}$. Then, define

@@

which depends on three additional parameters, $\theta_1$, $\theta_2$, and $x_{\max}$, such that $f(0)=\theta_1$ and $f(x_{\max})=\theta_4$. Under the parametric model (@eq-parametric), the squintability metric (@eq-squintability) can be shown to be equal to

@@

<!--To the best of our knowledge, -->

@@

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: Parameters estimated from the Gaussian process (including variance, range, smoothness, and nugget) and scaled logistic function ($\theta_1$ to $\theta_4$) for the pipe-finding and sine-wave finding problems.  The column "smooth" and "squint" represent the smoothness and squintability measures. 
dt <- tibble(shape = c(rep("pipe", 4), rep("sine", 8)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  mutate(index = ifelse(index == "dcor2d_2", "dcor2d", index)) 
dt |> 
  knitr::kable(
    digits = 2, format = "latex", 
    col.names = c(
      "shape", "index", "d",
      "variance", "range", "smooth", "nugget",
      "$\\theta_1$", "$\\theta_2$", "$\\theta_3$", "$\\theta_4$", "squint"), linesep = "",
    booktabs = T, escape = FALSE, row.names = TRUE) |> 
  column_spec(1, border_left = TRUE) |> 
  column_spec(c(4, 8, 13), border_right = TRUE) |> 
  kable_styling(font_size = 10) 
```

```{r}
#| label: tbl-mod-data
#| tbl-cap: The first 7 rows of the datasets processed for modelling. 
sim_df |> 
  dplyr::select(index, d, smoothness, 
    squintability, n_jellies, max_tries, 
    -I_max, P_J, time) |> 
  mutate(time = as.numeric(time)) |>
  arrange(index, d, n_jellies, max_tries) |> 
  head(7) |> 
  kable(digits = 2, format = "latex", 
        col.names = c("Index", "D", 
          "Smoothness", "Squintability", 
          "N. jellyfish", 
          "Max. tries", 
          "Success rate", "Time (sec)"),
        linesep = "",
        booktabs = T) |> 
  kable_styling(#latex_options = "hold_position",
    font_size = 10)
```

```{r}
#| label: tbl-mod-output
#| tbl-cap: Model estimates of proportion of jellyfish success on optimisation properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function. The variable smoothness, squintability, number of jellyfish and maximum number of tries are positively associated with JSO success rate while data dimension and being flagged as long runtime are negatively associated with the success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.
broom::tidy(mod1) |> 
  mutate(term = c("Intercept", "Smoothness",
   "Squintability", "Dimension (d)", 
   "Long time", "N. jellyfish", 
   "Max. tries")) |>
  kable(digits = 2, format = "latex", 
    linesep = "", booktabs = T) |> 
  column_spec(c(1,5), border_right = TRUE) |> 
  column_spec(1, border_left = TRUE) |> 
  kable_styling(font_size = 10) 
```

### 2024-06-30 optim.tex
lots of small typos fixed and to the best of our knowledge statement removed

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability and smoothness as introduced in
\citet{laa_using_2020}. Metrics for squintability and smoothness are
precisely defined here, something that is new for the field. A series of
simulation experiments using specially structured datasets and various
PP indexes are conducted to assess the behaviour of JSO under different
settings and its sensitivity to hyper-parameter choices. The
relationship between the JSO performance, hyper-parameter tuning and
properties of PP indexes is analysed to provide guidance for
practitioners using projection pursuit, and developers of new indexes.

@@

\begin{tcolorbox}[enhanced jigsaw, left=2mm, opacityback=0, toprule=.15mm, opacitybacktitle=0.6, coltitle=black, breakable, colframe=quarto-callout-note-color-frame, colback=white, arc=.35mm, bottomtitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toptitle=1mm, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, bottomrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]

@@

where \(\nu>0\) is the smoothness parameter and where
\(\mathcal{K}_\nu\) is the modified Bessel function. The Matérn
covariance function can be expressed analytically when \(\nu\) is a
half-integer, the most popular values in the literature being \(1/2\),
\(3/2\) and \(5/2\) . The parameter \(\nu\), called \emph{smoothness
parameter}, controls the decay of the covariance function. As such, it
is an appropriate measure of smoothness of a random field.

@@

which is a decreasing logistic function depending on two parameters
\(\theta_2\) and \(\theta_3\) , such that
\(\ell(\theta_{2})=\frac{1}{2}\). Then, define

@@

which depends on three additional parameters, \(\theta_1\), \(\theta_2\)
, and \(x_{\max}\), such that \(f(0)=\theta_1\) and
\(f(x_{\max})=\theta_4\). Under the parametric model
(\ref{eq-parametric}), the squintability metric (\ref{eq-squintability})
can be shown to be equal to

@@

To the best of our knowledge, this is the first attempt to measure
mathematically the notion of squintability.

@@

\centering\begingroup\fontsize{7}{9}\selectfont

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrrr>{}r|}
\toprule
  & shape & index & d & variance & range & smooth & nugget & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & squint\\
\midrule
1 & pipe & holes & 6 & 0.002 & 0.371 & 2.355 & 0.221 & 1.015 & 0.860 & 3.368 & 0.018 & 3.053\\
2 & pipe & holes & 8 & 0.000 & 0.178 & 2.189 & 0.820 & 1.014 & 0.869 & 3.264 & 0.029 & 2.960\\
3 & pipe & holes & 10 & 0.000 & 0.107 & 2.192 & 3.027 & 1.016 & 0.885 & 3.151 & 0.022 & 2.948\\
4 & pipe & holes & 12 & 0.000 & 0.148 & 2.295 & 1.581 & 1.011 & 0.878 & 3.345 & 0.004 & 3.117\\
5 & sine & MIC & 6 & 0.016 & 0.092 & 2.457 & 0.083 & 0.894 & 0.571 & 1.623 & -0.024 & 1.257\\
6 & sine & MIC & 8 & 0.016 & 0.081 & 2.636 & 0.084 & 0.932 & 0.328 & 1.314 & -0.030 & 0.772\\
7 & sine & TIC & 6 & 0.124 & 0.110 & 2.444 & 0.086 & 0.951 & 0.536 & 1.719 & -0.027 & 1.322\\
8 & sine & TIC & 8 & 0.125 & 0.099 & 2.475 & 0.085 & 0.949 & 0.564 & 1.723 & -0.028 & 1.369\\
9 & sine & dcor2d & 6 & 0.034 & 0.167 & 2.663 & 0.114 & 0.954 & 1.039 & 2.742 & -0.019 & 2.947\\
10 & sine & loess2d & 6 & 0.079 & 0.336 & 1.988 & 0.309 & 1.016 & 1.039 & 2.648 & 0.080 & 2.756\\
11 & sine & splines2d & 6 & 0.042 & 0.242 & 2.537 & 0.102 & 1.014 & 1.051 & 2.730 & -0.009 & 3.118\\
12 & sine & stringy & 6 & 0.000 & 0.007 & 1.536 & 38.166 & 1.045 & 0.011 & 254.734 & 0.053 & 2.957\\
\bottomrule
\end{tabular}
\endgroup{}

@@

\begingroup\fontsize{7}{9}\selectfont

\begin{longtable}[t]{lrrrrrrl}

@@

\tabularnewline

\toprule
Index & D & Smoothness & Squintability & Num. of jellyfish & Max. Num. of tries & Success rate & Time\\
\midrule
MIC & 6 & 2.457 & 1.257 & 20 & 50 & 0.12 & 2.479 secs\\
MIC & 6 & 2.457 & 1.257 & 20 & 100 & 0.24 & 8.950 secs\\
MIC & 6 & 2.457 & 1.257 & 50 & 50 & 0.52 & 5.651 secs\\
MIC & 6 & 2.457 & 1.257 & 50 & 100 & 0.64 & 13.223 secs\\
MIC & 6 & 2.457 & 1.257 & 100 & 50 & 0.76 & 19.453 secs\\
\addlinespace
MIC & 8 & 2.636 & 0.772 & 20 & 50 & 0.08 & 2.566 secs\\
MIC & 8 & 2.636 & 0.772 & 20 & 100 & 0.08 & 4.960 secs\\
\bottomrule

\end{longtable}

\endgroup{}

\begingroup\fontsize{7}{9}\selectfont

\begin{longtable}[t]{|>{}lrrr>{}r|}

@@

\tabularnewline

\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & -4.523 & 5.332 & -0.848 & 0.396\\
Smoothness & 1.195 & 1.915 & 0.624 & 0.533\\
Squintability & 2.060 & 0.684 & 3.011 & 0.003\\
dimension (d) & -0.628 & 0.255 & -2.462 & 0.014\\
long time & -0.874 & 1.286 & -0.680 & 0.497\\
\addlinespace
number of jellyfish & 0.216 & 0.127 & 1.701 & 0.089\\
maximum number of tries & 0.113 & 0.150 & 0.752 & 0.452\\
\bottomrule

\end{longtable}

\endgroup{}

>>>>

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on \citet{barnett1981interpreting}'s
squint angle) and smoothness as introduced in \citet{laa_using_2020}.
Here we mathematically define metrics for squintability and smoothness,
which is new for the field. A series of simulation experiments using
specially structured datasets and various PP indexes are conducted to
assess the behaviour of JSO under different settings and its sensitivity
to hyper-parameter choices. The relationship between the JSO
performance, hyper-parameter tuning and properties of PP indexes is
analysed to provide guidance for practitioners using projection pursuit,
and developers of new indexes.

@@

\begin{tcolorbox}[enhanced jigsaw, left=2mm, colback=white, opacityback=0, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, breakable, bottomrule=.15mm, titlerule=0mm, toprule=.15mm, opacitybacktitle=0.6]

@@

where \(\nu>0\) is the smoothness parameter and where
\(\mathcal{K}_\nu\) is the modified Bessel function. The Matérn
covariance function can be expressed analytically when \(\nu\) is a
half-integer, the most popular values in the literature being \(1/2\),
\(3/2\) and \(5/2\). The parameter \(\nu\), called \emph{smoothness
parameter}, controls the decay of the covariance function. As such, it
is an appropriate measure of smoothness of a random field.

@@

which is a decreasing logistic function depending on two parameters
\(\theta_2\) and \(\theta_3\), such that
\(\ell(\theta_{2})=\frac{1}{2}\). Then, define

@@

which depends on three additional parameters, \(\theta_1\),
\(\theta_2\), and \(x_{\max}\), such that \(f(0)=\theta_1\) and
\(f(x_{\max})=\theta_4\). Under the parametric model
(\ref{eq-parametric}), the squintability metric (\ref{eq-squintability})
can be shown to be equal to

@@

@@

\centering\begingroup\fontsize{10}{12}\selectfont

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrrr>{}r|}
\toprule
  & shape & index & d & variance & range & smooth & nugget & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & squint\\
\midrule
1 & pipe & holes & 6 & 0.00 & 0.37 & 2.36 & 0.22 & 1.02 & 0.86 & 3.37 & 0.02 & 3.05\\
2 & pipe & holes & 8 & 0.00 & 0.18 & 2.19 & 0.82 & 1.01 & 0.87 & 3.26 & 0.03 & 2.96\\
3 & pipe & holes & 10 & 0.00 & 0.11 & 2.19 & 3.03 & 1.02 & 0.88 & 3.15 & 0.02 & 2.95\\
4 & pipe & holes & 12 & 0.00 & 0.15 & 2.29 & 1.58 & 1.01 & 0.88 & 3.34 & 0.00 & 3.12\\
5 & sine & MIC & 6 & 0.02 & 0.09 & 2.46 & 0.08 & 0.89 & 0.57 & 1.62 & -0.02 & 1.26\\
6 & sine & MIC & 8 & 0.02 & 0.08 & 2.64 & 0.08 & 0.93 & 0.33 & 1.31 & -0.03 & 0.77\\
7 & sine & TIC & 6 & 0.12 & 0.11 & 2.44 & 0.09 & 0.95 & 0.54 & 1.72 & -0.03 & 1.32\\
8 & sine & TIC & 8 & 0.12 & 0.10 & 2.47 & 0.09 & 0.95 & 0.56 & 1.72 & -0.03 & 1.37\\
9 & sine & dcor2d & 6 & 0.03 & 0.17 & 2.66 & 0.11 & 0.95 & 1.04 & 2.74 & -0.02 & 2.95\\
10 & sine & loess2d & 6 & 0.08 & 0.34 & 1.99 & 0.31 & 1.02 & 1.04 & 2.65 & 0.08 & 2.76\\
11 & sine & splines2d & 6 & 0.04 & 0.24 & 2.54 & 0.10 & 1.01 & 1.05 & 2.73 & -0.01 & 3.12\\
12 & sine & stringy & 6 & 0.00 & 0.01 & 1.54 & 38.17 & 1.05 & 0.01 & 254.73 & 0.05 & 2.96\\
\bottomrule
\end{tabular}
\endgroup{}

@@

@@

\centering\begingroup\fontsize{10}{12}\selectfont

\begin{tabular}{lrrrrrrr}
\toprule
Index & D & Smoothness & Squintability & N. jellyfish & Max. tries & Success rate & Time (sec)\\
\midrule
MIC & 6 & 2.46 & 1.26 & 20 & 50 & 0.12 & 2.48\\
MIC & 6 & 2.46 & 1.26 & 20 & 100 & 0.24 & 8.95\\
MIC & 6 & 2.46 & 1.26 & 50 & 50 & 0.52 & 5.65\\
MIC & 6 & 2.46 & 1.26 & 50 & 100 & 0.64 & 13.22\\
MIC & 6 & 2.46 & 1.26 & 100 & 50 & 0.76 & 19.45\\
MIC & 8 & 2.64 & 0.77 & 20 & 50 & 0.08 & 2.57\\
MIC & 8 & 2.64 & 0.77 & 20 & 100 & 0.08 & 4.96\\
\bottomrule
\end{tabular}
\endgroup{}

@@

\centering\begingroup\fontsize{10}{12}\selectfont

\begin{tabular}{|>{}lrrr>{}r|}
\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & -4.52 & 5.33 & -0.85 & 0.40\\
Smoothness & 1.19 & 1.92 & 0.62 & 0.53\\
Squintability & 2.06 & 0.68 & 3.01 & 0.00\\
Dimension (d) & -0.63 & 0.26 & -2.46 & 0.01\\
Long time & -0.87 & 1.29 & -0.68 & 0.50\\
N. jellyfish & 0.22 & 0.13 & 1.70 & 0.09\\
Max. tries & 0.11 & 0.15 & 0.75 & 0.45\\
\bottomrule
\end{tabular}
\endgroup{}
